Reanalysis Step 2: Multiple Imputation

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
perfect prediction

::: {.callout-important}
## Key Change from Previous Version
`cohort_id` has been removed from the imputation model to avoid perfect prediction issues. The variable perfectly predicted outcome missingness (all 'excluded' had missing outcomes, all 'submitted' had observed outcomes), preventing successful imputation. The variable is preserved and added back after imputation for tracking purposes.
:::

@@

  # Temporal VA variables
  # NOTE: cohort_id REMOVED - it perfectly predicts outcome missingness
  "baseline_VA_logMAR",        # Baseline VA
  "intermediate_VA_logMAR",    # Intermediate follow-up VA
  "outcome_VA_logMAR",         # 2-year outcome
  # NOT outcome_va_vi_binary - derived from logMAR threshold

@@

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = dplyr::row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2),
                       labels = c("No_DR", "NPDR", "PDR"),
                       ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  ) |>
  # Fellow-eye variables (computed within patient clusters)
  dplyr::group_by(e_mrn_deidentified) |>
  dplyr::mutate(
    n_eyes = dplyr::n(),
    eye_index = dplyr::row_number(),
    fellow_baseline_VA = dplyr::if_else(n_eyes == 2, baseline_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_intermediate_VA = dplyr::if_else(n_eyes == 2, intermediate_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_outcome_VA = dplyr::if_else(n_eyes == 2, outcome_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_dr = dplyr::if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               factor(NA, levels = levels(person_dr), ordered = TRUE))
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-eye_index)
```

@@

# Temporal VA variables
custom_methods["baseline_VA_logMAR"] <- "pmm"
custom_methods["intermediate_VA_logMAR"] <- "pmm"
custom_methods["outcome_VA_logMAR"] <- "pmm"
# cohort_id removed from dataset to avoid perfect prediction

@@

# Never impute but can be used as predictors: ever_lapse_binary
if ("ever_lapse_binary" %in% rownames(pred_matrix)) pred_matrix["ever_lapse_binary", ] <- 0
# cohort_id removed entirely to avoid perfect prediction issues

@@

  # Note about cohort_id removal
  log_message("  - cohort_id removed from model to avoid perfect prediction")

@@

# Add back cohort_id and other variables not included in imputation
df_original_full <- df_for_imputation |>
  mutate(row_id = dplyr::row_number())

# Add back cohort_id from the original data
imputed_datasets <- lapply(imputed_datasets, function(imp_data) {
  # Add cohort_id back from original data based on row_id
  imp_data <- imp_data |>
    dplyr::left_join(
      df_original_full |> dplyr::select(row_id, cohort_id),
      by = "row_id"
    )
  imp_data
})

# Then add any other variables not included in imputation or transformations
vars_in_imputed <- names(imputed_datasets[[1]])
vars_to_add <- setdiff(names(df_original_full), c(vars_in_imputed, "row_id"))

>>>>

::: {.callout-important}
## Key Change from Previous Version
`cohort_id` is included in the dataset but NOT used as a predictor in the imputation model to avoid perfect prediction issues. The variable is set to never be imputed (method = "") and its column in the predictor matrix is set to 0 so it cannot predict other variables. This allows mice to understand the data structure while preventing perfect prediction problems.
:::

@@

  # Data indicator - will NOT be imputed or used as predictor
  "cohort_id",                 # Kept in dataset but not used for prediction

  # Temporal VA variables
  "baseline_VA_logMAR",        # Baseline VA
  "intermediate_VA_logMAR",    # Intermediate follow-up VA
  "outcome_VA_logMAR",         # 2-year outcome
  # NOT outcome_va_vi_binary - derived from logMAR threshold

@@

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = dplyr::row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2),
                       labels = c("No_DR", "NPDR", "PDR"),
                       ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),  # Keep as factor but won't impute
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  ) |>
  # Fellow-eye variables (computed within patient clusters)
  dplyr::group_by(e_mrn_deidentified) |>
  dplyr::mutate(
    n_eyes = dplyr::n(),
    eye_index = dplyr::row_number(),
    fellow_baseline_VA = dplyr::if_else(n_eyes == 2, baseline_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_intermediate_VA = dplyr::if_else(n_eyes == 2, intermediate_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_outcome_VA = dplyr::if_else(n_eyes == 2, outcome_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_dr = dplyr::if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               factor(NA, levels = levels(person_dr), ordered = TRUE))
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-eye_index)
```

@@

# Temporal VA variables
custom_methods["baseline_VA_logMAR"] <- "pmm"
custom_methods["intermediate_VA_logMAR"] <- "pmm"
custom_methods["outcome_VA_logMAR"] <- "pmm"

# Data indicator - never impute
custom_methods["cohort_id"] <- ""

@@

# Never impute these variables (set their rows to 0)
if ("ever_lapse_binary" %in% rownames(pred_matrix)) pred_matrix["ever_lapse_binary", ] <- 0
if ("cohort_id" %in% rownames(pred_matrix)) pred_matrix["cohort_id", ] <- 0

# CRITICAL: cohort_id must NOT be used as a predictor (set its column to 0)
if ("cohort_id" %in% colnames(pred_matrix)) pred_matrix[, "cohort_id"] <- 0

@@

  # Check cohort_id setup
  if ("cohort_id" %in% names(df_mice)) {
    log_message("  - cohort_id: included but NOT used as predictor")
    cohort_excluded <- sum(df_mice$cohort_id == "excluded")
    log_message(paste0("  - Records with cohort_id='excluded': ", cohort_excluded))
  }

@@

# Add back variables not included in imputation or transformations
df_original_full <- df_for_imputation |>
  mutate(row_id = dplyr::row_number())

# Add any other variables not included in imputation or transformations
vars_in_imputed <- names(imputed_datasets[[1]])
vars_to_add <- setdiff(names(df_original_full), c(vars_in_imputed, "row_id"))
